---
title: "Comparative Genomic Analysis Workflow"
author: "Megan Barkdull"
output: 
  github_document:
    toc: true
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This repository hosts the workflow for a comparative genomics analysis. The general overview is:

```{r, echo = FALSE, fig.align = 'center'}
library(DiagrammeR)
grViz("digraph flowchart {
      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle]        
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab7 [label = '@@7']
      tab8 [label = '@@8']
      tab9 [label = '@@9']
      tab10 [label = '@@10']
      tab11 [label = '@@11']
      tab12 [label = '@@12']
      tab13 [label = '@@13']
      
      # edge definitions with the node IDs
      tab1 -> tab2 -> tab3 -> tab4 -> tab5 -> tab6 -> tab7 -> tab8 -> tab13;
      tab5 -> tab9 -> tab10 -> tab13;
      tab3 -> tab11 -> tab12 -> tab13;
      }
      [1]: '1) Download the coding sequences, protein sequences, and annotation file for each species'
      [2]: '2) Filter for the longest isoform for each gene'
      [3]: '3) Clean gene names'
      [4]: '4) Translate coding sequences to amino acid sequences'
      [5]: '5) Run Orthofinder on amino acid sequences, generating orthogroups and multiple sequence alignments'
      [6]: '6) Prepare coding and amino acid sequences for PAL2NAL'
      [7]: '7) Generate nucleotide alignments with PAL2NAL'
      [8]: '8) Run BUSTED[S] on nucleotide alignments to test for positive selection'
      [9]: '9) Scale gene trees with Rubin pipeline'
      [10]: '10) Run RERConverge to correlate phenotypes and relative evolutionary rate'
      [11]: '11) Annotate coding sequences with InterProScan'
      [12]: '12) Concatenate InterProScan results'
      [13]: '13) Test for GO term enrichment with GOATOOLS'
      ")
```

### Using this workflow:
To use this workflow, simply clone this repository onto your own machine. You will run all Bash scripts in the main directory associated with this repository (i.e., in `/FormicidaeMolecularEvolution/`)

##  1. Downloading data:
For a project of this kind, you will need coding sequence and protein data, as well as GFF-formatted annotations, from your species of interest. The Bash script `DataDownload` can download these data for you. This script requires an input file, which should be a comma-delimited text file with:

  + Coding sequence download URLs in the first column
  + Protein sequence download URLs in the second column
  + GFF-formatted annotation download URLs in the third column
  + Species abbreviation codes in the fourth column; I suggest something like first letter of gene and first three letters of species. 
  + yes/no values in the fifth column; the indicates whether or not longest isoforms ought to be calculated for this species. See section 2 for details. 
  
See `./scripts/inputurls.txt` for an example. 

Once you have an input URLs file, simply run the script by executing the following command in a Bash shell:

`./scripts/DataDownload ./scripts/inputurls.txt`

This will create a new directory, `./1_RawData`, containing the downloaded input files. 

## 2. Selecting longest isoforms:
Protein and coding sequence datasets are likely to include multiple isoforms for many genes. This project makes use of just the single longest isoform for each gene. To filter down to just the longest isoforms, you'll use the script `GeneRetrieval`. This is an R script which uses the R package [orthologr](https://drostlab.github.io/orthologr/index.html) to filter for longest isoforms based on a protein sequence file and a GFF-formatted annotation file. The script then matches protein names to gene names in order to subset the coding sequence file to just the longest isoforms. 

Crucially, in order for this script to work properly, the script expects transcript gene names to look like this:
  
  + `>lcl|NW_012130065.1_cds_XP_012054525.1_1 [gene=LOC105617575] [db_xref=GeneID:105617575] [protein=proton-coupled amino acid transporter 4-like] [frame=2] [partial=5'] [protein_id=XP_012054525.1] [location=join(<125..532,740..1100,1848..2438)] [gbkey=CDS]`

And processes them to look like:
  
  + `XP_012054525.1`

Where `XP_012054525.1` is the name of a protein in the proteins file. 

To run this script, simply use the command:

`./scripts/GeneRetrieval /scripts/inputurls.txt`

This will create two files for each species, one with the longest protein isoforms and one with the longest coding sequence isoforms, in the directory `1_RawData`. 

**If you use this step, please cite orthologr as follows:**

Drost et al. 2015. Evidence for Active Maintenance of Phylotranscriptomic Hourglass Patterns in Animal and Plant Embryogenesis. Mol. Biol. Evol. 32 (5): 1221-1231. doi:10.1093/molbev/msv012

## 3. Cleaning the raw data:
The raw data is likely to have several features that will make future steps difficult or annoying. To solve this problem, you'll want to run the script `DataCleaning`. This will remove any special characters from gene names (for example, _ or (), which Orthofinder will change, causing errors) and add your four-letter taxon abbreviation to the beginning of each gene name. To run this script, simply use the command:

`./scripts/DataCleaning ./scripts/inputurls.txt`

This will create a new directory, `./CleanedData`, that contains the cleaned transcript files. 

## 3. Translating nucleotide sequences to amino acid sequences:
For many steps of this workflow, you'll actually need amino acid sequences rather than protein sequences. Therefore, we'll need to translate the data we downloaded. This will be done in one of two ways:

  + If the data you downloaded are raw transcript data, meaning for example that they don't start with a start codon, then we'll use Transdecoder to process and translate them into meaningful amino acid sequences. 
  + If they have already been processed and begin with start codons, we can take a simpler route and just translate them with the use of a codon table. 
  
The script `./scripts/DataTranslating` takes care of this process for us. It will attempt to run [Transdecoder](https://github.com/TransDecoder/TransDecoder/wiki) on each cleaned transcripts file; if the data is unprocessed, this will give us a translated output file. If the data is already processed, Transdecoder will fail to run and the script will instead translate the file with my Python script, `./scripts/TranscriptFilesTranslateScript.py`.

***This step uses the path to Transdecoder on Cornell's BioHPC. I plan to update the script in future so that it is more portabe. In the meantime, you'd need to edit lines 26 and 27 of `./scripts/DataTranslating` so that they point to Transdecoder on your machine.***

To run this step, simply use the command:

`./scripts/DataTranslating ./scripts/inputurls.txt`

This will create a new directory, `./TranslatedData/OutputFiles/`, that contains the translated transcript files. 

If you use Transdecoder, please cite it as:
Haas, B., and A. Papanicolaou. "TransDecoder." (2017).

## 4. Running Orthofinder:
Next, we will use [Orthofinder](https://davidemms.github.io) to identify groups of orthologous genes in our amino acid sequences, and to produce multiple sequence alignments with MAFFT. To run Orthofinder automatically, you can use the Bash script `DataOrthofinder`. ***This script takes a version number as a command line option to run either the most recent version ("current") or version 2.3.8. The script is currently written to work on Cornell's BioHPC, and so may require some changes to run on other machines (editing at least lines 14-16 and 42-46).***

To run this step, use the command:
"
`./scripts/DataOrthofinder [version number, either 'current' or '2.3.8'] [full path to a species tree where tip names correspond to file names, for example 'acep_transcripts']`

This will infer orthogroups and multiple-sequence alignments of amino acid sequences; MAFFT is used to generate the multiple sequence alignments. Outputs will be found in a new subdirectory of `/FormicidaeMolecularEvolution/`, `./OrthoFinder/fasta/OrthoFinder`.

If you use Orthofinder, please cite it:

  + OrthoFinder’s orthogroup and ortholog inference are described here:
      + Emms, D.M., Kelly, S. OrthoFinder: solving fundamental biases in whole genome comparisons dramatically improves orthogroup inference accuracy. Genome Biol 16, 157 (2015)
      + Emms, D.M., Kelly, S. OrthoFinder: phylogenetic orthology inference for comparative genomics. Genome Biol 20, 238 (2019)
  + If you use the OrthoFinder species tree then also cite:
      + Emms D.M. & Kelly S. STRIDE: Species Tree Root Inference from Gene Duplication Events (2017), Mol Biol Evol 34(12): 3267-3278
      + Emms D.M. & Kelly S. STAG: Species Tree Inference from All Genes (2018), bioRxiv https://doi.org/10.1101/267914

Please also cite MAFFT: 

  + K. Katoh, K. Misawa, K. Kuma, and T. Miyata. 2002. MAFFT: a novel method for rapid multiple sequence alignment based on fast Fourier transform. Nucleic Acids Res. 30(14): 3059-3066. 
  
## 5. Preparing files for PAL2NAL: 
### 5.1 Reordering multiple sequence alignment files for PAL2NAL
Orthofinder produces a single file for every individual orthogroup, containing the alignments for the sequences in that orthogroup. However, in order to create codon-aware alignments with PAL2NAL, we need files that contain all of the alignments for each individual species. 


```{r, echo = FALSE, fig.align = 'center'}
library(DiagrammeR)
grViz("digraph flowchart {
      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle]        
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab7 [label = '@@7']
      # edge definitions with the node IDs
      tab1 -> tab5;
      tab1 -> tab6;
      tab1 -> tab7;
      tab2 -> tab5;
      tab2 -> tab6;
      tab2 -> tab7;
      tab3 -> tab5;
      tab3 -> tab6;
      tab3 -> tab7;
      tab4 -> tab5;
      tab4 -> tab6;
      tab4 -> tab7;
      }
      [1]: 'Orthofinder file: \\n Locus1 Species1 \\n Locus1 Species2 \\n Locus1 Species3'
      [2]: 'Orthofinder file: \\n Locus2 Species1 \\n Locus2 Species2 \\n Locus2 Species3'
      [3]: 'Orthofinder file: \\n Locus3 Species1 \\n Locus3 Species2 \\n Locus3 Species3'
      [4]: 'Orthofinder file: \\n Locus4 Species1 \\n Locus4 Species2 \\n Locus4 Species3'
      [5]: 'Reshuffled file: \\n Locus1 Species1 \\n Locus2 Species1 \\n Locus3 Species1 \\n Locus4 Species1'
      [6]: 'Reshuffled file: \\n Locus1 Species2 \\n Locus2 Species2 \\n Locus3 Species2 \\n Locus4 Species2'
      [7]: 'Reshuffled file: \\n Locus1 Species3 \\n Locus2 Species3 \\n Locus3 Species3 \\n Locus4 Species3'
      ")
```


The R script `DataMSA.R` will recombine the Orthofinder outputs so that they can be input to PAL2NAL. To run the script, you'll need to get the full path to the MSA files created by Orthofinder in the previous step (it will be something like `/workdir/mb2337/FormicidaeMolecularEvolution/OrthoFinder/fasta/OrthoFinder/Results_Oct26/MultipleSequenceAlignments`). Then use the command:

`./scripts/DataMSA.R ./scripts/inputurls /FullPathToMSAFiles`

### 5.2 Filtering coding sequences for PAL2NAL:

The nucleotide sequence file that is run through PAL2NAL can have only the genes that are also present in the protein alignment file. Since our protein alignment files contain only a subset of genes, we need to filter through the nucleotide sequence files so that they too contain only this subset. To do this, use the R script `FilteringCDSbyMSA.R`. This script checks whether genes in the coding sequence file are present in the multiple sequence alignment file, then pulls the corresponding nucleotide sequences from the coding sequence file into a new, filtered output. 

To run this step, just use the command:
`./scripts/FilteringCDSbyMSA.R ./scripts/inputurls`

This script will output filtered coding sequence files to the subdirectory `./FilteredCDS`.

## 6. Generating codon-aware nucleotide alignments with PAL2NAL:
PAL2NAL will generate codon-aware nucleotide alignments, based on an input of an amino acid multiple sequence alignment and a nucleotide sequence. These inputs have been generated by the previous step. 

Now, to run PAL2NAL, use the Bash script `./scripts/DataRunPAL2NAL` by executing the command:

`./scripts/DataRunPAL2NAL ./scripts/inputurls`

This will produce a new directory, `./PAL2NALOutput`, containing an aligned nucleotide sequence file for each species. 

## 7. Testing for positive selection with BUSTED[S]

### 7.1 Assembling nucleotide sequence orthogroups for input to BUSTED[S]. 

To run BUSTED[S], we will need files that contain orthologous nucleotide sequences from each species. Therefore, we must recombine our codon-aware alignments in a step that is the inverse of step 5.1. To do this, use the R script `./scripts/DataSubsetCDS.R`. Run with the command:

`Rscript ./scripts/DataSubsetCDS.R ./scripts/inputurls_partial [path to multiple sequence alignments from Orthofinder]`

This will produce a new directory, `./CDSOrthogroups/MultipleSequenceAlignments/Output` that will contain files of orthologous nucleotide sequences for input to BUSTED[S].

### 7.2 Masking stop codons from orthogroup sequences
BUSTED[S] will not run on sequences which contain stop codons, even if these are reasonable, terminal stop codons. Hyphy includes a utility which will mask these these terminal stop codons in the orthogroups (there should be few-to-no other stop codons, because our alignments are codon-aware). To execute this step, use the script `/scripts/DataRemoveStopCodons`. 

### 7.3 Running BUSTED[S]

To run BUSTED[S], use the Bash script `./scripts/DataRunningBusted`. This script will run BUSTED[S] to test for positive selection on the sequences. 



### To do:

For the step where you must convert the outputs of OrthoFinder to be inputs for RERConverge, please see the [Comparative Genomics repository](https://github.com/mbarkdull/ComparativeGenomics)

